%
% Copyright © 2018 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%
%----------------------------------------------------------------------------------------
%\part{Lecture notes}
   \mychapter{Fields, units, and scales.}
      \input{qftLecture1.tex}
      \input{qftLecture2.tex}
      \section{Problems.}
         \input{qftLukeProblemSet1Problem4.tex}
         \input{ProblemSet1Problem3.tex}
   \mychapter{Lorentz transformations.}
      \input{qftLecture2LorentzTransformation.tex}
   %\mychapter{Lorentz transformations and a scalar action.}
      \input{qftLecture3.tex}
      \section{Problems.}
         \input{qftLukeProblemSet1Problem1.tex}
   \mychapter{Classical field theory.}
      \input{qftLecture3ClassicalFieldTheory.tex}
   %\mychapter{Scalar action, least action principle, Euler-Lagrange equations for a field}
      \input{qftLecture4.tex}
      \section{Problems.}
         \input{qftLecture3ClassicalFieldTheoryProblems.tex}
         \input{qftLukeProblemSet1Problem3.tex}
         \input{qftLukeProblemSet1Problem6.tex}
         \input{ProblemSet1Problem1.tex}
   \mychapter{Canonical quantization}
   %{Canonical quantization, Klein-Gordon equation, SHOs, momentum space representation, raising and lowering operators.}
      \input{qftLecture4CanonicalQuantization.tex}
      \input{qftLecture5.tex}
      \input{qftLecture6.tex}
      \section{Problems.}
         \input{scalarFieldCreationOpCommutator.tex}
         \input{scalarFieldHamiltonian.tex}
         \input{ProblemSet1Problem2.tex}
         \input{ProblemSet1Problem5.tex}
         \input{ProblemSet2Problem4.tex}
   %\mychapter{Symmetries, translation currents, energy momentum tensor.}
   \mychapter{Symmetries.}
      \input{qftLecture6Symmetries.tex}
      \input{qftLecture7.tex}
   %\mychapter{1st Noether theorem, spacetime translation current, energy momentum tensor, dilatation current.}
      \input{qftLecture8.tex}
   %\mychapter{Unbroken and spontaneously broken symmetries, Higgs Lagrangian, scale invariance, Lorentz invariance, angular momentum quantization.}
      \input{qftLecture9.tex}
      \section{Problems.}
         \input{noetherCurrentScalarField.tex}
         \input{qftLukeProblemSet1Problem5.tex}
         \input{ProblemSet1Problem4.tex}
         \input{ProblemSet2Problem2.tex}
         \input{ProblemSet2Problem3.tex}
\ifthenelse{\boolean{redacted}}
{
\relax
}
{
         \input{ProblemSet5Problem2.tex}
}
   \mychapter{Lorentz boosts, generators, Lorentz invariance, microcausality.}
      \input{qftLecture10.tex}
   %\mychapter{Microcausality, Lorentz invariant measure
      \input{qftLecture11.tex}
   \mychapter{External sources.}
      \input{qftLecture11ExternalSources.tex}
   %\mychapter{Klein-Gordon Green's function, Feynman propagator path deformation, Wightman function, Retarded Green's function.}
      \input{qftLecture12.tex}
      %\section{Problems.}
      %   \input{ProblemSet2Problem1.tex}
   %\mychapter{Forced Klein-Gordon equation.}
      \input{qftLecture13.tex}
   %\mychapter{Coherent states, Number density.}
      \input{qftLecture13CoherentStates.tex}
      \section{Problems.}
         \input{ProblemSet2Problem1.tex}
         \input{qftLukeProblemSet1Problem2.tex}
   %\mychapter{Time ordered product, perturbation theory, Heisenberg picture, interaction picture, Dyson's formula.}
   \mychapter{Perturbation theory.}
      \input{qftLecture13FeynmansGreenFunction.tex}
      %   \input{nonhomoKGhamiltonian.tex}
   %\mychapter{Time evolution, Hamiltonian pertubation, ground state.}
      \input{qftLecture14.tex}
   %\mychapter{Perturbation ground state, time evolution operator, time ordered product, interaction.}
      \input{qftLecture15.tex}
      \input{qftLecture15b.tex}
      \section{Problems.}
         \input{nonhomoKGhamiltonian.tex}
         \input{ProblemSet4Problem1.tex}
         \input{qftLecture15Problems.tex}
         \input{braOmega.tex}
         \input{ProblemSet3Problem1.tex}
         \input{ProblemSet3Problem2.tex}
         \input{ProblemSet3Problem4.tex}
   \mychapter{Scattering and decay.}
      \input{qftLecture15bScattering.tex}
   %\mychapter{Differential cross section, scattering, pair production, transition amplitude, decay rate, S-matrix, connected and amputated diagrams, vacuum fluctuation, symmetry coefficient.}
      \input{qftLecture16.tex}
   %\mychapter{Scattering, decay, cross sections in a scalar theory.}
      \input{qftLecture17.tex}
      \input{qftLecture18.tex}
      \section{Problems.}
         \input{ProblemSet4Problem2.tex}
         \input{ProblemSet4Problem3.tex}
         \input{ProblemSet3Problem3.tex}
   \mychapter{Fermions, and spinors.}
      \input{qftLecture19.tex}
      \input{qftLecture20.tex}
      \input{qftLecture21.tex}
      \input{qftLecture21b.tex}
      \input{qftLecture22.tex}
      \input{qftLecture23.tex}
      \section{Problems.}
         \input{ProblemSet4Problem4.tex}
         \input{qftLecture19Problems.tex}
         \input{qftLecture20Problems.tex}
         \input{squarerootpsigma.tex}
         \input{qftLecture21aProblems.tex}
         \input{qftLecture21bProblems.tex}
         \input{qftLecture23Problems.tex}
