%
% Copyright © 2018 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%
\makeproblem{ Scale invariance and conserved charge }{qft:problemSet1:4}{
Consider classical electrodynamics with the Lagrangian
\begin{dmath}\label{eqn:ProblemSet1Problem4:20}
S = \int d^4 x \lr{ -\inv{4} F_{\mu\nu} F^{\mu\nu} }.
\end{dmath}
Consider the following ``dilatation'' (or ``scale'') transformation:
\begin{dmath}\label{eqn:ProblemSet1Problem4:40}
\begin{aligned}
x_\mu &\rightarrow x'_\mu = e^d x_\mu \\
A_\mu(x) &\rightarrow A'_\mu(x') = e^{-d} A_\mu(x),
\end{aligned}
\end{dmath}
where \( d \) is a constant, called the dilatation parameter.

Dilatation invariance in QED (and QCD) is perhaps the simplest example of a symmetry, where the classical action is invariant, but the quantum theory is not (as you will learn later, in the spring class). Broken scale invariance arises because one has to introduce a short-distance cutoff (a UV ``regulator'') to define the quantum theory. (We already saw an indication of the need for a regulator when we considered the divergent zero point energy of the free quantum scalar field.)

\makesubproblem{}{qft:problemSet1:4a}
Show that the action is invariant under dilatations.
\makesubproblem{}{qft:problemSet1:4b}
Find the corresponding Noether current.
\makesubproblem{}{qft:problemSet1:4c}
Show that -- perhaps, after a redefinition of \( j_\mu \) ; notice that any conserved current \( j_\mu \) can be
redefined by adding to it \( \partial^\nu C_{\mu\nu} \), where \( C_{\mu\nu} \) is antisymmetric, without spoiling its conservation
(in this case \( C \) can depend on \( x^\mu, \partial^\mu \) and \( A^\mu \), of course) the dilatation current is simply related
to the energy-momentum tensor: \( j^{\text{con f}}_\mu = x_\nu {{T^\nu}_\mu}^{\text{con f}}\), where the symbol con f indicates that
these are the conformal energy-momentum tensor and dilatation current. Notice that this problem, secretly, requires you to also derive \( T^{\mu\nu} \) for the electromagnetic field.
\makesubproblem{}{qft:problemSet1:4d}
Show, then, that conservation of \(  j^{\text{con f}}_\mu  \) implies that the energy-momentum tensor of classical
electrodynamics is traceless (the trace of the tensor is defined as usual to be \( g_{\mu\nu} T^{\mu\nu}\)).
\makesubproblem{}{qft:problemSet1:4e}
Finally, open your classical electrodynamics books and recall the interpretation of the \( T^{00}, T^{xx},T^{yy} \), etc., components of the energy momentum tensor as energy density and pressure. Show that the tracelessness of \( T^{\mu\nu} \) is equivalent to the familiar relation
\( p = \rho/3 \) between the energy density and pressure of isotropic radiation -- the equation of state of blackbody radiation.

} % makeproblem

\makeanswer{qft:problemSet1:4}{
\makeSubAnswer{}{qft:problemSet1:4a}
With \( {x'}^\mu = e^{d} x^\mu \), the volume element transforms as
\begin{dmath}\label{eqn:ProblemSet1Problem4:60}
d^4 x' \rightarrow e^{4d} d^4 x.
\end{dmath}
The components of the four-gradient transform as
\begin{dmath}\label{eqn:ProblemSet1Problem4:80}
\PD{x'_\mu}{}
=
\PD{x'_\mu}{x_\mu}
\PD{x_\mu}{}
=
e^{-d}
\PD{x_\mu}{},
\end{dmath}
so
\begin{dmath}\label{eqn:ProblemSet1Problem4:100}
F'_{\mu\nu} =
\partial'_{\mu} A'_\nu
-
\partial'_{\nu} A'_\mu
=
e^{-2d} F_{\mu\nu}.
\end{dmath}
The action is therefore invariant
\begin{dmath}\label{eqn:ProblemSet1Problem4:120}
S'
= -\inv{4} \int d^4 x' F'_{\mu\nu} {F'}^{\mu\nu}
= -\inv{4} \int e^{4d} d^4 x e^{2d} F_{\mu\nu} e^{2d} F^{\mu\nu}
= -\inv{4} \int d^4 x F_{\mu\nu} F^{\mu\nu}
= S.
\end{dmath}
\makeSubAnswer{}{qft:problemSet1:4a}
We need the variation of the potential
\begin{dmath}\label{eqn:ProblemSet1Problem4:180}
\delta A_{\nu}
= A'_{\nu}(x) - A_{\nu}(x)
= A'_{\nu}(e^{-d} x') - A_{\nu}(x)
\approx A'_{\nu}((1 -d) x') - A_{\nu}(x)
= e^{-d} A_{\nu}((1 -d) x') - A_{\nu}(x)
\approx (1-d) \lr{ A_{\nu} - d x^\alpha \partial_\alpha A_{\nu}} - A_{\nu}
=
- d x^\alpha \partial_\alpha A_{\nu}
- d \lr{ A_{\nu} - d x^\alpha \partial_\alpha A_{\nu}}
\approx
-d(1+ x^\alpha \partial_\alpha )A_{\nu},
\end{dmath}
and the variation of the field
\begin{dmath}\label{eqn:ProblemSet1Problem4:140}
\delta F_{\mu\nu}
= F'_{\mu\nu}(x) - F_{\mu\nu}(x)
= F'_{\mu\nu}(e^{-d} x') - F_{\mu\nu}(x)
\approx F'_{\mu\nu}((1 -d) x') - F_{\mu\nu}(x)
= e^{-2d} F_{\mu\nu}((1 -d) x') - F_{\mu\nu}(x)
\approx (1-2d) \lr{ F_{\mu\nu} - d x^\alpha \partial_\alpha F_{\mu\nu}} - F_{\mu\nu}
=
- d x^\alpha \partial_\alpha F_{\mu\nu}
-2 d \lr{ F_{\mu\nu} - d x^\alpha \partial_\alpha F_{\mu\nu}}
\approx
-d(2+ x^\alpha \partial_\alpha )
F_{\mu\nu},
\end{dmath}
so the variation of the Lagrangian is
\begin{dmath}\label{eqn:ProblemSet1Problem4:160}
\delta \LL
=
-\inv{2} (\delta F_{\mu\nu}) F^{\mu\nu}
=
-\inv{2}
(-d)
F^{\mu\nu}
(2+ x^\alpha \partial_\alpha )F_{\mu\nu}
=
(d)
F^{\mu\nu}
F_{\mu\nu}
+
\frac{d}{2}
F^{\mu\nu} x^\alpha \partial_\alpha F_{\mu\nu}
=
(d)
F^{\mu\nu}
F_{\mu\nu}
+
\frac{d}{4}
x^\alpha \partial_\alpha \lr{ F_{\mu\nu} F^{\mu\nu} }
=
-4 (d) \LL - (d) x^\alpha \partial_\alpha \LL
=
-4 (d) \LL - (d) \lr{ \partial_\alpha (x^\alpha \LL) - \LL \partial_\alpha x^\alpha }
=
- (d) \partial_\alpha (x^\alpha \LL),
\end{dmath}
so the variational current (what is this called?) is
\begin{dmath}\label{eqn:ProblemSet1Problem4:200}
J^\mu_d
=
- (d) x^\mu \LL.
\end{dmath}
Finally, we need
\begin{dmath}\label{eqn:ProblemSet1Problem4:220}
\PD{(\partial_\mu A_\nu)}{\LL}
=
-\inv{2}
F^{\alpha\beta}
\PD{(\partial_\mu A_\nu)}{}\lr{ \partial_\alpha A_\beta - \partial_\beta A_\alpha}
=
-\inv{2}
\lr{
   F^{\mu\nu}
   -
   F^{\nu\mu}
}
=
-F^{\mu\nu}.
\end{dmath}
Combining \cref{eqn:ProblemSet1Problem4:220}, \cref{eqn:ProblemSet1Problem4:200}, and \cref{eqn:ProblemSet1Problem4:180} we can calculate the conserved current, which is (for \( d = 1 \)) is
\begin{dmath}\label{eqn:ProblemSet1Problem4:240}
j^\mu
=
\PD{(\partial_\mu A_\nu)}{\LL} \delta A_\nu - J^\mu_d
=
F^{\mu\nu}
\lr{
   A_\nu + x^\alpha \partial_\alpha A_{\nu}
}
+
x^\mu \LL.
\end{dmath}

\makeSubAnswer{}{qft:problemSet1:4b}
TODO.
\makeSubAnswer{}{qft:problemSet1:4c}
TODO.
\makeSubAnswer{}{qft:problemSet1:4d}
TODO.
\makeSubAnswer{}{qft:problemSet1:4e}
TODO.
}
