%
% Copyright © 2017 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%
\input{../latex/blogpost.tex}
\renewcommand{\basename}{qftLecture8}
\renewcommand{\dirname}{notes/phy2403/}
\newcommand{\keywords}{PHY2403H}
\input{../latex/peeter_prologue_print2.tex}

%\usepackage{phy2403}
\usepackage{peeters_braket}
\usepackage{peeters_layout_exercise}
\usepackage{peeters_figures}
\usepackage{mathtools}
\usepackage{siunitx}
\usepackage{macros_cal} % LL

\beginArtNoToc
\generatetitle{PHY2403H Quantum Field Theory.  Lecture 8: 1st Noether theorem, spacetime translation current, energy momentum tensor, dilatation current.  Taught by Prof.\ Erich Poppitz}
%\chapter{1st Noether theorem, dilatation current}
\label{chap:qftLecture8}

\paragraph{DISCLAIMER: Very rough notes from class, with some additional side notes.}

These are notes for the UofT course PHY2403H, Quantum Field Theory I, taught by Prof. Erich Poppitz fall 2018.
%, covering \textchapref{{1}} \citep{peskin1995introduction} content.

\section{1st Noether theorem.}

Recall that, given a transformation
\begin{dmath}\label{eqn:qftLecture8:20}
\phi(x) \rightarrow \phi(x) + \delta \phi(x),
\end{dmath}
such that the transformation of the Lagrangian is only changed by a total derivative
\begin{dmath}\label{eqn:qftLecture8:40}
\LL(\phi, \partial_\mu \phi) \rightarrow
\LL(\phi, \partial_\mu \phi)
+ \partial_\mu J_\epsilon^\mu,
\end{dmath}
then there is a conserved current
\begin{dmath}\label{eqn:qftLecture8:60}
j^\mu = \PD{(\partial_\mu \phi)}{\LL} \delta_\epsilon \phi - J_\epsilon^\mu.
\end{dmath}
Here \( \epsilon \) is an x-independent quantity (i.e. a \underline{global symmetry}).
This is in contrast to ``gauge symmetries'', which can be more accurately be categorized as a redundancy in the description.

As an example, give
\begin{dmath}\label{eqn:qftLecture8:80}
\phi(x) \rightarrow \phi(x) - a^\mu \partial_\mu \phi
\end{dmath}
\begin{dmath}\label{eqn:qftLecture8:100}
\LL(\phi, \partial_\mu \phi) \rightarrow
\LL(\phi, \partial_\mu \phi)
- a^\mu \partial_\mu \LL
\LL(\phi, \partial_\mu \phi)
+ \partial_\mu \lr{ -{\delta^\mu}_\nu a^\nu \partial_\mu \LL }
\end{dmath}
Here \( J^\mu_\epsilon = \evalbar{J^\mu_\epsilon}{\epsilon = a^\nu} \), and the current is
\begin{dmath}\label{eqn:qftLecture8:120}
J^\mu = (\partial^\mu \phi)(-a^\nu \partial_\nu \phi) + {\delta^{\mu}}_\nu a^\nu \LL.
\end{dmath}
In particular, we have one such current for each \( \nu \), and we write
\begin{dmath}\label{eqn:qftLecture8:140}
{T^\mu}_\nu = -(\partial^\mu \phi)(\partial_\nu \phi) + {\delta^{\mu}}_\nu \LL.
\end{dmath}
By Noether's theorem, we must have
\begin{dmath}\label{eqn:qftLecture8:160}
\partial_\mu
{T^\mu}_\nu = 0, \quad \forall \nu.
\end{dmath}
Fixme: check this!

\paragraph{Example: our potential Lagrangian}

\begin{dmath}\label{eqn:qftLecture8:180}
\LL = \inv{2} \partial^\mu \phi \partial_\nu \phi - \frac{m^2}{2} \phi^2 - \frac{\lambda}{4} \lambda^4
\end{dmath}
Written with upper indexes
\begin{dmath}\label{eqn:qftLecture8:200}
T^{\mu\nu}
= -(\partial^\mu \phi)(\partial^\nu \phi) + g^{\mu\nu} \LL
= -(\partial^\mu \phi)(\partial^\nu \phi) + g^{\mu\nu} \lr{
\inv{2} \partial^\alpha \phi \partial_\nu \phi - \frac{m^2}{2} \phi^2 - \frac{\lambda}{4} \lambda^4
}
\end{dmath}

There are 4 conserved currents \( J^{\mu(\nu)} = T^{\mu\nu} \).  Observe that this is symmetric (\( T^{\mu\nu} = T^{\nu\mu} \)).

We have four associated charges
\begin{dmath}\label{eqn:qftLecture8:220}
Q^\nu = \int d^3 x T^{0 \nu}.
\end{dmath}
We call
\begin{dmath}\label{eqn:qftLecture8:240}
Q^0 = \int d^3 x T^{0 0},
\end{dmath}
the energy density, and call
\begin{dmath}\label{eqn:qftLecture8:260}
P^i = \int d^3 x T^{0 i},
\end{dmath}
(i = 1,2,3) the momentum density.

writing this out explicitly the energy density is
\begin{dmath}\label{eqn:qftLecture8:280}
T^{00}
= - \dot{\phi}^2 + \inv{2} \lr{ \dot{\phi}^2 - (\spacegrad \phi)^2 - \frac{m^2}{2} - \frac{\lambda}{4} \phi^4}
= -\lr{
\inv{2} \dot{\phi}^2 + \inv{2} (\spacegrad \phi)^2 + \frac{m^2}{2} + \frac{\lambda}{4} \phi^4
},
\end{dmath}
and
\begin{dmath}\label{eqn:qftLecture8:300}
T^{0i} = \partial^0 \partial^i \phi,
\end{dmath}
\begin{dmath}\label{eqn:qftLecture8:320}
P^{i} = -\int d^3 x\partial^0 \phi \partial^i \phi
\end{dmath}
Since the energy density is negative definite (due to an arbitrary choice of translation sign), let's redefine \( T^{\mu\nu} \) to have a positive sign
\begin{dmath}\label{eqn:qftLecture8:340}
T^{00}
\equiv
\inv{2} \dot{\phi}^2 + \inv{2} (\spacegrad \phi)^2 + \frac{m^2}{2} + \frac{\lambda}{4} \phi^4,
\end{dmath}
and
\begin{dmath}\label{eqn:qftLecture8:360}
P^{i} = \int d^3 x\partial^0 \phi \partial^i \phi
\end{dmath}

As an operator we have
\begin{dmath}\label{eqn:qftLecture8:380}
\hatQ = \int d^3 x \hatT^{00} =
\int d^3 x
\lr{
\inv{2} \pihat^2 + \inv{2} (\spacegrad \phihat)^2 + \frac{m^2}{2} \phihat^2 + \frac{\lambda}{4} \phihat^4
}.
\end{dmath}
\begin{dmath}\label{eqn:qftLecture8:400}
\hatP^{i} = \int d^3 x \pihat \partial^i \phi
\end{dmath}

We showed that
\begin{dmath}\label{eqn:qftLecture8:420}
\ddt \hatO = i \antisymmetric{\hatH}{\hatO}
\end{dmath}
This implied that \( \phihat, \pihat \) obey the classical EOMs
\begin{equation}\label{eqn:qftLecture8:440}
\ddt \phihat = i \antisymmetric{\hatH}{\phihat} = \ddt{\pihat}
\end{equation}
\begin{equation}\label{eqn:qftLecture8:460}
\ddt \pihat = i \antisymmetric{\hatH}{\pihat} = ...
\end{equation}

In terms of creation and annihilation operators (for the \( \lambda = 0 \) free field), up to a constant
\begin{dmath}\label{eqn:qftLecture8:480}
\hatH
= \int d^3 x \hatT^{00}
= \int \frac{d^3 p}{(2 \pi)^3} \omega_\Bp \hata_\Bp^\dagger \hata_\Bp
\end{dmath}
Fixme: show that:
\begin{dmath}\label{eqn:qftLecture8:500}
\hatP^i
= \int d^3 x \pihat \partial^i \phihat
= \cdots
= \int \frac{d^3 p}{(2 \pi)^3} p^i \hata_\Bp^\dagger \hata_\Bp
\end{dmath}
Now we see the energy and momentum as conserved quantities associated with spacetime translation.

\section{Unitary operators}

In QM we say that \( \hat{\Bp} \) ``generates translations''.

With \( \hat{\Bp} \equiv -i \Hbar \spacegrad \) that translation is
\begin{equation}\label{eqn:qftLecture8:520}
\hatU = e^{i \Ba \cdot \hat{\Bp}} = e^{\Ba \cdot \spacegrad}
\end{equation}

In particular
\begin{dmath}\label{eqn:qftLecture8:540}
\bra{\Bx} \hatU \ket{\psi} = e^{\Ba \cdot \hat{\Bp} } \psi(\Bx) = \psi(\Bx + \Ba).
\end{dmath}

In one dimension
\begin{dmath}\label{eqn:qftLecture8:560}
\hatU \hat{x} \hatU^\dagger
=
e^{\Ba \cdot \hat{p} } \psi(\Bx)
e^{-\Ba \cdot \hat{p} }
= \hat{\Bx} + a \hat{\BOne}.
\end{dmath}
This uses the Baker-Campbell-Hausdorff formula.
\maketheorem{ Baker-Campbell-Hausdorff }{thm:qftLecture8:580}{
\begin{dmath}\label{eqn:qftLecture8:600}
e^{B} A e^{-B} = \sum_{n = 0}^\infty \inv{n!} \antisymmetric{B \cdots}{\antisymmetric{B}{A}},
\end{dmath}
where the n-th commutator is denoted above

\begin{itemize}
\item \( n = 1 \) : \( \antisymmetric{B}{A} \)
\item \( n = 2 \) : \( \antisymmetric{B}{\antisymmetric{B}{A}} \)
\item \( n = 3 \) : \( \antisymmetric{B}{\antisymmetric{B}{\antisymmetric{B}{A}}} \)
\end{itemize}
} % theorem

Proof:

\begin{dmath}\label{eqn:qftLecture8:620}
f(t) = e^{tB} A e^{-tB}
= f(0) + t f'(0) + \frac{t^2}{2} f''(0) + \cdots \frac{t^n}{n!} f^{(n)}(0)
\end{dmath}

\begin{dmath}\label{eqn:qftLecture8:640}
f(0) = A
\end{dmath}
\begin{dmath}\label{eqn:qftLecture8:660}
f'(t)
=
e^{tB} B A e^{-tB}
+
e^{tB} A (-B) e^{-tB}
=
e^{tB} \antisymmetric{B}{A} e^{-tB}
\end{dmath}
\begin{dmath}\label{eqn:qftLecture8:680}
f''(t) =
e^{tB} B \antisymmetric{B}{A} e^{-tB}
+
e^{tB} \antisymmetric{B}{A} (-B) e^{-tB}
=
e^{tB} \antisymmetric{B}{\antisymmetric{B}{A}} e^{-tB}.
\end{dmath}
From
\begin{dmath}\label{eqn:qftLecture8:700}
f(1)
= f(0) + f'(0) + \inv{2} f''(0) + \cdots \inv{n!} f^{(n)}(0)
\end{dmath}
we have
\begin{dmath}\label{eqn:qftLecture8:720}
e^{B} A e^{-B} = A +
\antisymmetric{B}{A} + \inv{2} \antisymmetric{B}{\antisymmetric{B}{A}} + \cdots
\end{dmath}

Example:
\begin{dmath}\label{eqn:qftLecture8:740}
e^{a \partial_x} x e^{-a \partial_x } = x + a \antisymmetric{\partial_x}{x} + \cdots
= x + a.
\end{dmath}

\paragraph{Application:}

\begin{dmath}\label{eqn:qftLecture8:760}
e^{i \text{Hermitian} } = \text{unitary}
\end{dmath}
\begin{dmath}\label{eqn:qftLecture8:860}
e^{i \text{Hermitian} } \times
e^{-i \text{Hermitian} }
= 1
\end{dmath}
So
\begin{dmath}\label{eqn:qftLecture8:780}
\hatU(\Ba) =
e^{i a^j \hat{p}^j }
\end{dmath}
is a unitary operator representing finite translations in a Hilbert space.

\begin{dmath}\label{eqn:qftLecture8:800}
\hatU(\Ba) \phihat(\Bx) \hatU^\dagger(\Ba)
=
e^{i a^j \hat{p}^j }
\phihat(\Bx)
e^{-i a^k \hat{p}^k }
=
\phihat(\Bx)
+ i a^j \antisymmetric{\hatP^j}{\phihat(\Bx)} + \frac{-a^{j_1} a^{j_2}}{2} \antisymmetric{\hatP^{j_1}}{\antisymmetric{\hatP^{j_2}}{\phihat(\Bx)}}
\end{dmath}

\begin{dmath}\label{eqn:qftLecture8:820}
\antisymmetric{\hatP^j}{\phihat(\Bx)}
=
\int d^3 y \antisymmetric{\pihat(\By) \partial^j \phihat(\By)}{\phihat(\Bx)}
=
\int d^3 y \antisymmetric{\pihat(\By)}{\phihat(\Bx} \partial^j \phihat(\By)
=
\int d^3 y (-i ) \delta^3(\By - \Bx) \partial^j \phihat(\By)
=
-i \partial^j \phihat(\Bx).
\end{dmath}

\begin{dmath}\label{eqn:qftLecture8:840}
\hatU(\Ba) \phihat(\Bx) \hatU^\dagger(\Ba)
= \phihat(\Bx) + i a^j (-i) \partial^j \phihat(\Bx) + \cdots
= \phihat(\Bx) + a^j \partial^j \phihat(\Bx) + \cdots
= \phihat(\Bx + \Ba)
\end{dmath}

\section{Continuous symmetries}

For all infinitesimal transformations, continuous symmetries lead to conserved charges \( Q \).  In QFT we map these charges to Hermitian operators \( Q \rightarrow \hatQ \).  We say that these charges are ``generators of the corresponding symmetry'' through unitary operators
\begin{dmath}\label{eqn:qftLecture8:880}
\hatU = e^{i \text{parameter} \hatQ}.
\end{dmath}
These represent the action of the symmetry in the Hilbert space.

\paragraph{Example: spatial translation}
\begin{dmath}\label{eqn:qftLecture8:900}
\hatU(\Ba) = e^{i \Ba \cdot \hat{\BP}}
\end{dmath}
\paragraph{Example: time translation}
\begin{dmath}\label{eqn:qftLecture8:920}
\hatU(t) = e^{i t \hat{H}}.
\end{dmath}

\section{Classical scalar theory}

For \( d > 2 \) let's look at
\begin{dmath}\label{eqn:qftLecture8:940}
S =
\int d^d x \lr{
\inv{2} \partial
\inv{2} \partial^\mu \phi \partial_\nu \phi - \frac{m^2}{2} \phi^2 - \lambda \phi^{d-2}
}
\end{dmath}
Take \( m^2, \lambda \rightarrow 0 \), the free massless scalar field.
We have a shift symmetry in this case since \( \phi(x) \rightarrow \phi(x) + \text{constant} \).
The current is just
\begin{dmath}\label{eqn:qftLecture8:960}
j^\mu
= \PD{(\partial_\mu \phi)}{\phi} \delta \phi - \cancel{J^\mu}
= \text{constant} \times \partial^\mu \phi
= \partial^\mu \phi,
\end{dmath}
where the constant factor has been set to one.
This current is clearly conserved since \( \partial_\mu J^\mu = \partial_\mu \partial^\mu \phi = 0\) (the equation of motion).

These are called ``Goldstein bosons''.

With \( m \rightarrow 0, \lambda \ne 0 \) we have
\begin{dmath}\label{eqn:qftLecture8:980}
S =
\int d^d x \lr{
\inv{2} \partial
\inv{2} \partial^\mu \phi \partial_\nu \phi - \tilde{\lambda} \phi^{d-2}
}
\end{dmath}
Here we have a scale or dilatation invariance
\begin{dmath}\label{eqn:qftLecture8:1000}
x \rightarrow e^{\lambda} x
\end{dmath}
\begin{dmath}\label{eqn:qftLecture8:1020}
\phi(x) \rightarrow \phi'(x') = e^{-(d-2) \lambda/2} \phi
\end{dmath}

\begin{dmath}\label{eqn:qftLecture8:1040}
d^3 x = e^{d\lambda} d^3 x
\end{dmath}

\begin{dmath}\label{eqn:qftLecture8:1060}
(\partial_\mu \phi)^2 \rightarrow e^{-(d-2)\lambda} e^{-2 \lambda} \lr{ \partial_\mu \phi(x) }^2
\end{dmath}

\begin{dmath}\label{eqn:qftLecture8:1080}
\phi^{2 d/(d -2)} \rightarrow \lr{\phi'(x')}^{2d/(d-2)}
=
e^{-(d-2)/2 \times 2d/(d-2) \lambda} \lr{ \phi }^{2d/(d-2)}
\end{dmath}

(switching to \(d = 4\))
to find Noether currrent ...

\begin{dmath}\label{eqn:qftLecture8:1100}
\delta_\lambda \phi = \phi'(x) - \phi(x)
= - \lambda(1 + x^\mu \partial_\mu ) \phi
\end{dmath}

In free field with \( \tilde{\lambda} = 0 \)

\begin{dmath}\label{eqn:qftLecture8:1120}
\delta_\lambda \lr{ \inv{2} \partial_\mu \phi \partial^\mu \phi}
= \partial^\mu \delta_\lambda (\partial_\mu \phi)
= \partial^\mu \phi \partial_\mu ( - \lambda( 1 + x^\nu \partial_\nu )\phi )
= -\lambda \lr{ \partial^\mu \phi \partial_\mu \phi + \partial^\mu \phi \partial_\mu \lr{ x^\nu \partial_\nu \phi } }
= -\lambda \lr{ 2 \partial^\mu \phi \partial_\mu \phi + \partial^\mu \phi x^\nu \partial_\nu \phi }
=
\end{dmath}

Wrong.  Starting over.

We need the transformation of the derivatives of \( \phi \).

\begin{dmath}\label{eqn:qftLecture8:1140}
\inv{2} \partial^\mu \phi \partial_\mu \phi
\rightarrow
\inv{2} {\partial'}^\mu \phi' {\partial'}_\mu \phi'
=
\inv{2} e^{-4\lambda}
\partial^\mu \phi \partial_\mu \phi
\end{dmath}
Let
\begin{dmath}\label{eqn:qftLecture8:1160}
\begin{aligned}
\Phi'(x')
&=
{\partial'}^\mu \phi' {\partial'}_\mu \phi' \\
\Phi(x) &=
\partial^\mu \phi \partial_\mu \phi,
\end{aligned}
\end{dmath}

so
\begin{dmath}\label{eqn:qftLecture8:1180}
\begin{aligned}
\Phi'(x') &= e^{-4 \lambda} \Phi(x) \\
\Phi'(e^{\lambda} x) &= e^{-4 \lambda} \Phi(x)
\end{aligned}
\end{dmath}
...
so

\begin{dmath}\label{eqn:qftLecture8:1200}
\delta \LL = - \lambda \lr{ 4 + x^\mu \partial_\mu } \lr{
\partial^\mu \phi \partial_\mu \phi
}
=
-\inv{2} \lambda \partial_\mu \lr{ x^\mu \partial_\nu \phi \partial^\nu \phi}
=
-\lambda \lr{ \partial_\mu x^\mu
\inv{2} (\partial_\nu \phi)^2 + x^\mu \partial_\mu \lr{ \inv{2} \partial_\nu \phi \partial^\nu \phi }
}
\end{dmath}

so
\begin{dmath}\label{eqn:qftLecture8:1220}
\begin{aligned}
J^\mu_\lambda &= \frac{-x^\mu}{2} \partial_\nu \phi \partial^\nu \phi \\
\delta \phi &= - \lr{ 1 + x^\mu \partial_\mu } \phi
\end{aligned}
\end{dmath}

\begin{dmath}\label{eqn:qftLecture8:1240}
j^\mu = \PD{(\partial_\mu \phi)}{\LL} \delta \phi - J^\mu
\end{dmath}

In the end, find
\begin{dmath}\label{eqn:qftLecture8:1260}
j^\mu = -\partial^\mu \phi \lr{ 1 + x^\nu \partial_\nu } \phi + \inv{2} x^\mu \partial_nu \phi \partial^\nu \phi
\end{dmath}

Flipping signs
\begin{dmath}\label{eqn:qftLecture8:1280}
j^\mu_{\text{dil}} = \partial^\mu \phi \lr{ 1 + x^\nu \partial_\nu } \phi - \inv{2} x^\mu \partial_nu \phi \partial^\nu \phi
\end{dmath}
or
\begin{dmath}\label{eqn:qftLecture8:1300}
j^\mu_{\text{dil}} = -x_\nu T^{\nu \mu} - \inv{2} \partial^\mu \phi^2
\end{dmath}

The current and \( T^{\mu\nu} \) can both be redefined \( j^{\mu'} = j^\mu + \partial_\nu C^{\nu\mu} \) adding an antisymetric \( C^{\mu\nu} = -C^{\nu\mu} \)

\begin{dmath}\label{eqn:qftLecture8:1320}
j^\mu_{\text{dil conformal}} = - x_\nu T^{\nu\mu}_{\text{conformal}}
\end{dmath}

\begin{dmath}\label{eqn:qftLecture8:1340}
\partial_\mu
j^\mu_{\text{dil conformal}} = - {{T_{\text{conformal}}}^\mu}_\mu
\end{dmath}

%}
%\EndArticle
\EndNoBibArticle
