%
% Copyright © 2017 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%
%{
\input{../latex/blogpost.tex}
\renewcommand{\basename}{qftLecture14}
\renewcommand{\dirname}{notes/phy2403/}
\newcommand{\keywords}{PHY2403H}
\input{../latex/peeter_prologue_print2.tex}

%\usepackage{phy2403}
\usepackage{peeters_braket}
%\usepackage{peeters_layout_exercise}
\usepackage{peeters_figures}
\usepackage{mathtools}
\usepackage{siunitx}
\usepackage{macros_cal} % LL

\newcommand{\ultensor}[3]{{{#1}^{#2}}_{#3}}

\beginArtNoToc
\generatetitle{PHY2403H Quantum Field Theory.  Lecture 14: XXX.  Taught by Prof.\ Erich Poppitz}
%\chapter{XXX}
\label{chap:qftLecture14}

%s/_{\text{I,int}}/_{\\text{I,int}}

%%Peeter's lecture notes from class.  These may be incoherent and rough.
%%
%%These are notes for the UofT course PHY2403H, Quantum Field Theory, taught by Prof. Erich Poppitz, covering \textchapref{{1}} \citep{peskin1995introduction} content.

\paragraph{DISCLAIMER: Very rough notes from class, with some additional side notes.}

These are notes for the UofT course PHY2403H, Quantum Field Theory, taught by Prof. Erich Poppitz, fall 2018.
%, covering \textchapref{{1}} \citep{peskin1995introduction} content.

\section{Review}

Given a field \( \phi(t_0, \Bx) \), satisfying the commutation relations
\begin{dmath}\label{eqn:qftLecture14:20}
\antisymmetric{\pi(t_0, \Bx)}{\phi(t_0, \By)} = -i \delta(\Bx - \By)
\end{dmath}
we introduced an interaction picture field given by
\begin{dmath}\label{eqn:qftLecture14:40}
\phi_I(t, x) = e^{i H_0(t- t_0)} \phi(t_0, \Bx) e^{-iH_0(t - t_0)}
\end{dmath}
related to the Heisenburg picture representation by
\begin{dmath}\label{eqn:qftLecture14:60}
\phi_H(t, x)
= e^{i H(t- t_0)} \phi(t_0, \Bx) e^{-iH(t - t_0)}
= U^\dagger(t, t_0) \phi_I(t, \Bx) U(t, t_0),
\end{dmath}
where \( U(t, t_0) \) is the time evolution operator.
\begin{dmath}\label{eqn:qftLecture14:80}
U(t, t_0) =
e^{i H_0(t - t_0)}
e^{i H(t - t_0)}
\end{dmath}
We argued that
\begin{dmath}\label{eqn:qftLecture14:100}
i \PD{t}{} U(t, t_0) = H_{\text{I,int}}(t) U(t, t_0)
\end{dmath}
We found the glorious expression
\begin{dmath}\label{eqn:qftLecture14:120}
U(t, t_0)
= T \exp{\lr{ -i \int_{t_0}^t H_{\text{I,int}}(t') dt'}}
=
\sum_{n = 0}^\infty \frac{(-i)^n}{n!} \int_{t_0}^t dt_1 dt_2 \cdots dt_n T\lr{ H_{\text{I,int}}(t_1) H_{\text{I,int}}(t_2) \cdots H_{\text{I,int}}(t_n) }
\end{dmath}

However, what we are really after is
\begin{dmath}\label{eqn:qftLecture14:140}
\bra{\Omega} T(\phi(x_1) \cdots \phi(x_n)) \ket{\Omega}
\end{dmath}
Such a product has many labels and names, and we'll describe it as ``vacuum expectation values of time-ordered products of arbitrary \#s of local Heisenberg operators''.

\section{Pertubation}

\begin{dmath}\label{eqn:qftLecture14:n}
\begin{aligned}
H &= \text{exact Hamiltonian} = H_0 + H_int
H_0 &= \text{free Hamiltonian.  We know all about \( H_0 \) and assume that it has a lowest (groundstate) \( \ket{0} \), the ``vacuum'' state of \( H_0 \).}
\end{aligned}
\end{dmath}

H has eigenstates, in particular, assumed to have a unique ground state \( \ket{\Omega} \) satisfying
\begin{dmath}\label{eqn:qftLecture14:n}
H \ket{\Omega}  = \ket{\Omega} E_0,
\end{dmath}
and has states \( \ket{n} \), representing excited (non-vacuum states with energies > \( E_0 \)).
These states are assumed to be a complete basis
\begin{dmath}\label{eqn:qftLecture14:n}
\BOne = \ket{\Omega}\bra{\Omega} + \sum_n \ket{n}\bra{n} + \int dn \ket{n}\bra{n}.
\end{dmath}
The latter is sometimes written with a superimposed sum-integral notation: F1.

For some time \( T \) we have
\begin{dmath}\label{eqn:qftLecture14:n}
e^{-i H T} \ket{0} = e^{-i H T} 
\lr{
   \ket{\Omega}\braket{\Omega}{0} + \sum_n \ket{n}\braket{n}{0} + \int dn \ket{n}\braket{n}{0}.
},
\end{dmath}
and now pull a fast one, considering a formal transformation \( T \rightarrow T(1 - i \epsilon) \), where \( \epsilon \rightarrow 0^+ \), and consider very large \( T \).  This gives
\begin{dmath}\label{eqn:qftLecture14:n}
\lim_{T \rightarrow \infty, \epsilon \rightarrow 0^+} 
e^{-i H T(1 - i \epsilon)} \ket{0} 
= 
\lim_{T \rightarrow \infty, \epsilon \rightarrow 0^+} 
e^{-i H T(1 - i \epsilon)} 
\lr{
   \ket{\Omega}\braket{\Omega}{0} + \sum_n \ket{n}\braket{n}{0} + \int dn \ket{n}\braket{n}{0}.
}
= 
\lim_{T \rightarrow \infty, \epsilon \rightarrow 0^+} 
e^{-i E_0 T - E_0 T \epsilon} 
   \ket{\Omega}\braket{\Omega}{0} + \sumint_n e^{-i E_n T - \epsilon E_n T} \ket{n}\braket{n}{0}
=
\lim_{T \rightarrow \infty, \epsilon \rightarrow 0^+} 
e^{-i E_0 T - E_0 T \epsilon} 
\lr{
   \ket{\Omega}\braket{\Omega}{0} + \sumint_n e^{-i (E_n -E_0) T - \epsilon T (E_n - E_0)} \ket{n}\braket{n}{0}
}
\end{dmath}
The limits are evaluated by first taking \( T \) to infinity, then only after that take \( \epsilon \rightarrow 0^+ \).  Doing this, the sum is dominated by the ground state contribution, since each excited state also has a \( e^{-\epsilon T(E_n - E_0)} \) suppression factor (in addition to the leading suppression factor).

Less formally, we can write
\begin{dmath}\label{eqn:qftLecture14:n}
\sumint \ket{n}\bra{n} \rightarrow 
\sum_k \int \frac{d^3 p}{(2 \pi)^3} \ket{\Bp, k}\bra{\Bp, k}
\end{dmath}
where \( k \) is some unknown quantity that we are summing over.
If we have 
\begin{dmath}\label{eqn:qftLecture14:n}
H \ket{\Bp, k} = E_{\Bp, k} \ket{\Bp, k},
\end{dmath}
then
\begin{dmath}\label{eqn:qftLecture14:n}
e^{-i H T} \sumint \ket{n}\bra{n}
=
\sum_k \int \frac{d^3 p}{(2 \pi)^3} \ket{\Bp, k} e^{-i E_{\Bp, k} \bra{\Bp, k}.
\end{dmath}
If we take matrix elements
\begin{dmath}\label{eqn:qftLecture14:n}
\bra{A} 
e^{-i H T} \sumint \ket{n}\bra{n} \ket{B}
=
\sum_k \int \frac{d^3 p}{(2 \pi)^3} \braket{A}{\Bp, k} e^{-i E_{\Bp, k} \braket{\Bp, k}{B}
=
\sum_k \int \frac{d^3 p}{(2 \pi)^3} e^{-i E_{\Bp, k} f(\Bp).
\end{dmath}
If we assume that \( f(\Bp) \) is a well behaved smooth function, we have ``infinite'' frequency oscillation within a windowed function, as sketched in
F2
The Riemann-Lebescue lemma \citep{wiki:RiemannLebesgue} describes such integrals, the result of which is that such an integral goes to zero.  This is a different hand waving argument to argue that we can examine only the ground state contribution above.
\begin{dmath}\label{eqn:qftLecture14:n}
e^{-i H T} \ket{0} = e^{-i E_0 T} \ket{\Omega}\braket{\Omega}{0}
\end{dmath}
in the \( T \rightarrow \infty(1 - i \epsilon) \) limit.  We can now write the ground state as

\begin{dmath}\label{eqn:qftLecture14:n}
\ket{\Omega}
= 
\evalbar{
\frac{ e^{i E_0 T - i H T } \ket{0} }{
\braket{\Omega}{0} 
}
}{ T \rightarrow \infty(1 - i \epsilon) }
=
\evalbar{
\frac{ e^{i H (T - t_0) } \ket{0} }{
e{-i E_0(T + t_0)} \braket{\Omega}{0} 
}
}{ T \rightarrow \infty(1 - i \epsilon) }
=
\evalbar{
\frac{ e^{i H (t_0 - (-T)) } e^{ -i H_0 (-T - t_0) } \ket{0} }{
e{-i E_0(t_0 - (-T)} \braket{\Omega}{0} 
}
}{ T \rightarrow \infty(1 - i \epsilon) }
\end{dmath}
where we use \( H_0 \ket{0} = 0 \).

With 
\begin{dmath}\label{eqn:qftLecture14:n}
U(t, t_0)
= e^{i H_0(t - t_0)} e^{-i H(t - t_0)}
= T \exp{\lr{ -i \int_{t_0}^t H_{\text{I,int}}(t') dt'}}
\end{dmath}

\paragraph{Claim:} Similarly (DIY) 
\begin{dmath}\label{eqn:qftLecture14:n}
U(t, t')
= e^{i H_0(t - t_0)} e^{-i H(t - t')} e^{-i H_ (t' - t_0)}
= T \exp{\lr{ -i \int_{t'}^t H_{\text{I,int}}(t'') dt''}}
\end{dmath}
Observe that we recover \cref{eqn:qftLecture14:120} when \( t' = t_0 \).

This gives
\boxedEquation{eqn:qftLecture14:n}{
\ket{\Omega}
= \frac{U(t_0, -T) \ket{0} }{e^{-i E_0(t_0 - (-T))} \braket{\Omega}{0}
}.

We are close to where we want to be.  Wednesday we will start scattering and Feynman diagrams.

%}
\EndArticle
