%
% Copyright © 2018 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%

%
%\chapter{Preface}
% this suppresses an explicit chapter number for the preface.
\chapter*{Preface}%\normalsize
  \thispagestyle{empty}
  \addcontentsline{toc}{chapter}{Preface}

This is a set of course notes from the Fall 2018, University of Toronto
Quantum Field Theorry (PHY2403), taught by Prof.\ Erich Poppitz.

\paragraph{The course syllabus included the following topics:}
\begin{itemize}
\item Introduction: Energy and distance scales; units and conventions. Uncertainty relations in the relativistic domain and the need for multiple particle description.
\item Canonical quantization. Free scalar field theory.
\item Symmetries and conservation laws.
\item Interacting fields: Feynman diagrams and the S matrix; decay widths and phase space.
\item Spin \(1/2\) fields: Spinor representations, Dirac and Weyl spinors, Dirac equation. Quantizing fermi fields and statistics.
\item Vector fields and Quantum electrodynamics. 
\end{itemize}

\paragraph{This book contains:}

\begin{itemize}
\item Lecture notes.
\item Personal notes exploring auxiliary details.
\item Worked practice problems.
\ifthenelse{\boolean{redacted}}%
{%
%\item Links to Matlab, Mathematica, and Julia notebooks associated with the course material and problems (but not problem sets).
}%
{
\item My solutions (as-is, with errors.) for problem sets 1-4.%
}
\ifthenelse{\boolean{with-mathematica}}
{
\item Links to Mathematica notebooks associated with problems and course material.%
}
{
}
\end{itemize}

\paragraph{On the problem set solutions:}
These notes are no longer redacted and include whatever portions of the problem set 1-4 solutions I completed, errors and all.  In the event that any of the problem sets are recycled for future iterations of the course, students who are taking the course (all mature grad students pursuing science for the love of it, not for grades) are expected to act responsibly, and produce their own solutions, within the constraints provided by the professor.

\paragraph{Thanks:}
My thanks go to Professor Poppitz for teaching this course, to the study gang, and to
Emily Tyhurst and Stefan Divic who kindly provided me their notes for lecture 22.

%
Peeter Joot  \quad peeterjoot@pm.me
